---
title: "Expected credit loss - explanatory data analysis"
author: "Eligijus Bujokas"
date: "12/19/2020"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(data.table)
library(ggplot2)
library(zoo)
library(stringr)
library(gridExtra)
library(reshape)
Sys.setlocale("LC_ALL","C")
```

# Overview

This document is used to explore the data of the company called "LendingClub". It is an American company that provides short term loans both for individuals and for bussineses. 

To know more about the company, visit: https://www.lendingclub.com/ 

All the files and codes used in this project are stored in the repo: https://github.com/Eligijus112/expected-credit-loss 

The data can be found in kaggle: https://www.kaggle.com/denychaen/lending-club-loans-rejects-data

The data has information about all approved loans from 2007 up until 2019 3th quarter. 

# Reading the data 

```{r "Reading data"}
# Using fread to read the data
d <- fread('data/Loan_status_2007-2020Q3.gzip')
  
# Total number of loans in the data set
print(nrow(d))

# Total number of columns 
print(ncol(d))
```

In the dataset there are 150 features regarding a unique loan. We will analyze and use in models only some of the features:

**addr_state** - The state provided by the borrower in the loan application

**annual_inc** - The self-reported annual income provided by the borrower during registration.

**dti** - debt to income ratio; a ratio calculated using the borrower’s total monthly debt payments on the total debt obligations, excluding mortgage and the requested LC loan, divided by the borrower’s self-reported monthly income.

**emp_length** - Employment length in years. Possible values are between 0 and 10 where 0 means less than one year and 10 means ten or more years. 

**earliest_cr_line** - the month at which the earliest credit line was opened for the borrower.

**funded_amnt** - The total amount committed to that loan at that point in time.

**home_ownership** - The home ownership status provided by the borrower during registration or obtained from the credit report. Our values are: RENT, OWN, MORTGAGE, OTHER.

**loan_amnt** - The listed amount of the loan applied for by the borrower. If at some point in time, the credit department reduces the loan amount, then it will be reflected in this value.

**loan_status** - current status of the loan.

**pub_rec** - Number of derogatory public records.

**pub_rec_bankruptcies** - Number of public record bankruptcies

**purpose** - A category provided by the borrower for the loan request. 

**issue_d** - The date at which the loan was issued.

**term** - number of payments on the loan. Values can be either 36 or 60.

**int_rate** - interest rate on the loan.

**recoveries** - post charge off gross recovery.

**total_rec_prncp** - principal received to date.

**verification_status** - Indicates if income was verified by LC, not verified, or if the income source was verified.

**delinq_2yrs** - The number of 30+ days past-due incidences of delinquency in the borrower's credit file for the past 2 years.


```{r "Important columns"}
# Defining all the columns used in the analysis
cols_to_use = c(
  "addr_state",
  "annual_inc",
  'emp_length',
  'earliest_cr_line',
  'funded_amnt',
  'home_ownership',
  'loan_amnt',
  'loan_status',
  'pub_rec',
  'pub_rec_bankruptcies',
  'purpose',
  'issue_d',
  'term',
  'int_rate',
  'recoveries',
  'total_rec_prncp',
  'dti',
  'verification_status',
  "delinq_2yrs",
  "initial_list_status",
  "hardship_flag",
  "grade"
)

# Subseting the data
d <- d[, ..cols_to_use]

# Printing the top of the data table
head(d)
```
# Data cleaning and engineering

Before analyzing the various splits of the data there are some columns that need to be cleaned and engineered. 

## Date columns 

```{r dates}
# Creating the splited columns 
d[, c("issue_m", "issue_y"):=tstrsplit(issue_d, "-")]

# Getting the last two digits of issue_y
d[, issue_y:=str_sub(issue_y, -2)]

# Creating the formated column
d[, issue_d_formated:=zoo::as.yearmon(paste(issue_m, issue_y, sep="-"), "%b-%y")]

# Creating the splited columns 
d[, c("cr_m", "cr_y"):=tstrsplit(earliest_cr_line, "-")]

# Getting the last two digits of issue_y
d[, cr_y:=str_sub(cr_y, -2)]

# Wrangling the earliest cr column
d[, earliest_cr_line_formated:=zoo::as.yearmon(paste(cr_m, cr_y, sep="-"), "%b-%y")]

# Calculating the difference between the two dates
d[, day_diff:=difftime(issue_d_formated, earliest_cr_line_formated, units='days')]

# Converting to numeric
d[, day_diff:=as.numeric(day_diff)]

# Dropping negative or missing values from the dataframe; these are very old accounts and compise a fraction of percent of the whole data
d <- d[d$day_diff > 0 & !is.na(d$day_diff)]

# Calculating the month diff 
d[, month_diff:=round(day_diff/30)]
```

## Historic loan issues 

```{r loansbytime}
# Aggregating by issue date and sorting
loan_n = d[, .N, by="issue_d_formated"] %>% 
  .[order(issue_d_formated)] %>% 
  .[, issue_year:=year(issue_d_formated)] %>%
  .[, N_y:=sum(N), by='issue_year'] %>% 
  setnames(., c('N', 'N_y'), c('loans_monthly', 'loans_yearly'))

# Ploting the time series
p1 <- ggplot(data=loan_n, aes(x=issue_d_formated, y=loans_monthly)) +
  geom_bar(stat="identity", fill="steelblue", color='black') + 
  ggtitle('Monthly count of issued loans')

p1
```

## Creating the Y variable 

```{r y variable}
# Printing out unique loan statuses
print(unique(d$loan_status))

# Function to encode the good loans as 1 and 0 as bad loans
create_Ypd <- function(x, bad_loans=c('Late (31-120 days)', "Default", "Charged Off", "Does not meet the credit policy. Status:Charged Off")){
  return(ifelse(is.element(x, bad_loans), 'bad', 'good'))
}

# Creating the Y variable
d[, Y:=create_Ypd(loan_status)]

# Saving the variable as factor
d$Y <- as.factor(d$Y)
```

```{r subset}
d <- d[issue_d_formated < zoo::as.yearmon('2019-01')]

print(nrow(d))

# Visualizing the distribution of Y 
bymonth <- d[, .N, by=c("issue_d_formated", "Y")] %>% 
  cast(., issue_d_formated ~ Y, value="N") %>% 
  data.table(.) %>% 
  .[, rate:=bad * 100/(bad + good)]

# Calculating the shares
ggplot(data=bymonth, aes(x=issue_d_formated, y=rate)) + geom_area(color="darkblue", fill="lightblue")
```

Issued loans by grade

```{r}
share_by_grade = d[, .("total_issued_grade" = .N), by=c('issue_d_formated', 'grade')]
total_issued = d[, .("total_issued" = .N), by='issue_d_formated']

# Merging the information 
merged = merge(share_by_grade, total_issued, all.x=TRUE)

# Calculating the share
merged[, share:=total_issued_grade/total_issued]

ggplot(data=merged, aes(x=issue_d_formated, y=share, fill=grade)) + geom_bar(position="stack", stat="identity", color='black')
```

## Summary statistics of numeric features

```{r summary}
# Cleaning the interest rates
d$int_rate <- as.numeric(gsub("%", "", d$int_rate))

summary(d[, c('annual_inc', 'loan_amnt', 'pub_rec', 'int_rate', 'dti', 'month_diff')])
```


## Debt to income ratio

```{r dti}
ggplot(data=d[dti<100], aes(x=dti, fill=Y)) + geom_density(alpha=0.5)

# Droping the rows with negative dti 
d <- d[dti>=0]
```

## Annual income

```{r annual_income}
ggplot(data=d[annual_inc < 200000], aes(x=annual_inc, fill=Y)) + geom_density(alpha=0.5)
```

## Loan amount

```{r loan_amount}
ggplot(data=d, aes(x=loan_amnt, fill=Y)) + geom_density(alpha=0.5)
```

## Interest rate 

```{r interest rate}
ggplot(data=d, aes(x=int_rate, fill=Y)) + geom_density(alpha=0.5)
```

## Cleaninng the employment length variable

```{r emp length}
clean_emp_length <- function(x){
  return(ifelse(x=="", "< 1 year", x))
}

d[, emp_length:=clean_emp_length(emp_length)]
```

## Cleaaning the hadsdhip flag variable

```{r hardship}
clean_hardship <- function(x){
  if(x == ''){
    return('missing')
  } else {
    return(x)
  }
}

d$hardship_flag <- sapply(d$hardship_flag, clean_hardship)
```


# LGD and EAD variables

The data for modeling is all charged off loans. 

```{r charged_off}
loan_default_data = d[loan_status %in% c("Charged Off", "Does not meet the credit policy. Status:Charged Off")]
```

## LGD independant variable - recovery rate

```{r recovery rate, warning=FALSE}
loan_default_data[, recovery_rate:=recoveries/funded_amnt]
loan_default_data[, recovery_rate:=ifelse(recovery_rate>1, 1, recovery_rate)]

# Ploting the recovery rate distribution
p1 <- ggplot(data=loan_default_data, aes(x=recovery_rate)) + geom_histogram(fill='steelblue', color='black') + ggtitle('Full data')

p1
```


## EAD independant variable - credit conversion factor

```{r credit conversion factor, warning=FALSE}
loan_default_data[, CCF:=(funded_amnt - total_rec_prncp)/funded_amnt]

# Ploting the recovery rate distribution
p1 <- ggplot(data=loan_default_data, aes(x=CCF)) + geom_histogram(fill='steelblue', color='black') + ggtitle('Full data')

p1
```

# Creating the datasets for further modeling 

## Spliting to train and test sets

```{r}
train <- d[issue_d_formated < zoo::as.yearmon('2018-01')]
test <- d[issue_d_formated >= zoo::as.yearmon('2018-01')]
```

## Weight of evidence

To infer what categorical feature values are the most influential in determining bad loan from a good loan we can use the weight of evidence (WOE for short) criteria. For a feature $i$ and the feature level $j$ the $WOE_{i, j}$ is calculated with the following formula:

$$WOE_{i, j} = log \left(\dfrac{P(X_{i}=j|Y=1)}{P(X_{i} = j|Y=0)}\right)$$
Lets define a function in R that calculates the WOE statistic for any given feature. Note, that the dependent variable needs to be called **Y**.

```{r woe}
woe <- function(feature, data){
  # Grouping by the feature and getting the total number of observations in each level
  woeTable <- data[, .N, by=c("Y", feature)] %>% 
    # Seting the names 
    setnames(., c('Y', 'feature', 'N')) %>% 
    # Casting so that Y is in the columns
    cast(., feature ~ Y, value="N") %>% 
    # Converting back to data.table
    data.table(.) %>% 
    # Filling the missing values with 0 
    setnafill(., fill=0, cols=c('good', 'bad')) %>%
    # Getting the proportions of good and bad borrowers in each level of the variable
    .[, c('prop_good', 'prop_bad') := list(good/sum(good), bad/sum(bad))] %>% 
    # Calculating the WOE statistic 
    .[, woe:=log(prop_good/prop_bad)] %>% 
    # Sorting by the woe stat
    .[order(woe, decreasing=TRUE)]
  
  return(woeTable)
}

# Example of the function output:
woeResult = woe('purpose', train)
print(woeResult)

# Defining a function to plot the woe 
plotWoe <- function(ft, woeData){
  p <- ggplot(woeData, aes(x=reorder(feature, woe), y=woe)) + 
            geom_bar(stat="identity", fill="steelblue", color='black') + 
            ggtitle(ft) + 
            labs(x='Feature value') +
            coord_flip()
   return(p)
}
```

## Iterating through all categorical features 

The initial weight of evidence plots:

```{r}
cat_vars <- c("purpose", "emp_length", "grade", "home_ownership", 'initial_list_status', 'verification_status')
plots <- list()

for(cat_var in cat_vars){
  woeStat <- woe(cat_var, train)
  plots[[cat_var]] <- plotWoe(cat_var, woeStat)
}

grid.arrange(plots$purpose, plots$emp_length, plots$grade, plots$home_ownership, plots$initial_list_status, plots$verification_status, ncol = 3, nrow=3)
```

```{r adjusting the addr_state variable}
adj_addr_state <- function(x){
  if (is.element(x, c('DC', 'VT', 'NH'))){
    return("DC:VT:NH")
  } else if (is.element(x, c('ME', 'CO', 'WV', 'WY', 'WA', 'SC', 'KS'))){
    return("ME:CO:WV:WY:WA:SC:KS")
  } else if (is.element(x, c('ND', 'NM', 'SD', 'TN', 'LA', 'NV'))){
    return("ND:NM:SD:TN:LA:NV")
  } else if (is.element(x, c("OK", 'AL', 'AR', 'MS', 'NE'))){
    return('OK:AL:AR:MS:NE')
  } else {
    return("other")
  }
}

train$addr_state_groups <- sapply(train$addr_state, adj_addr_state)

woeStat <- woe('addr_state_groups', train)
p <- plotWoe('addr_state_groups', woeStat)
p
```

```{r adjusting the purpose variable}
adj_purpose <- function(x){
  if (is.element(x, c('wedding', 'car'))){
    return("wedding:car")
  } else if (is.element(x, c('credit_card', 'home_improvement'))){
    return("credit_card:home_improvemenet")
  } else if (is.element(x, c('small_business'))){
    return("small_business")
  } else if (is.element(x, c('renewable_energy', "educational", "moving"))){
    return('renewable_energy:educ:moving')
  } else {
    return("other")
  }
}

train$purpose_groups <- sapply(train$purpose, adj_purpose)

woeStat <- woe('purpose_groups', train)
p <- plotWoe('addr_state_groups', woeStat)
p
```

As we can see from the plots, the only adjustment will be needed to emp_length variable. 

```{r adjust emp_length}
adj_emp_length <- function(x){
  if(is.element(x, ("< 1 year"))){
    return("< 1 year")
  } else if (x=='1 year') {
    return ('1 year')
  } else if (is.element(x, c('2 years', '3 years', '4 years'))){
    return ('2-4 years')
  } else if (is.element(x, c('5 years', '6 years', '7 years')))(
    return('5-7 years')
  ) else if (is.element(x, c('8 years', '9 years'))){
    return('8-9 years') 
  } else {
    return('10+ years')
  }
}

train$emp_length_groups <- sapply(train$emp_length, adj_emp_length)

woeStat <- woe('emp_length_groups', train)
p <- plotWoe('emp_length_groups', woeStat)
p
```

## Annual income

```{r}
capvalues <- function(x, cap){
  return(min(c(x, cap)))
}

plotWoeNumeric <- function(ft, woeData){
  p <- ggplot(woeData, aes(x=feature, y=woe)) + 
            geom_bar(stat="identity", fill="steelblue", color='black') + 
            ggtitle(ft) + 
            labs(x='Feature value') +
            coord_flip()
   return(p)
}

# Droping the outliers
train$annual_inc_c <- sapply(train$annual_inc, capvalues, cap=300000)

# Ploting the distribution
ggplot(data=train, aes(x=annual_inc_c)) + geom_density(fill='blue')

# Cutting the annual income into 10 equal sized partitions
train$annual_inc_groups = cut(train$annual_inc_c / 1000, 10)

woedf <- woe('annual_inc_groups', train)
plotWoeNumeric('annual_income', woedf)
```

Defining a function for annual income split. 

```{r final_income}
adj_income <- function(x){
  if(x < 30000){
    return("[0,30)")
  } else if (x < 60000){
    return("[30,60)")
  } else if (x < 100000){
    return("[60,100)")
  } else if (x < 200000){
    return("[100,200)")
  } else {
    return("200+")
  }
}

train$annual_inc_groups <- sapply(train$annual_inc, adj_income)

woedf <- woe('annual_inc_groups', train)
plotWoe('Final annual income splits', woedf)
```

## Loan amount

```{r loan_amnt}
# Cutting the annual income into 10 equal sized partitions
train$loan_amnt_groups = cut(train$loan_amnt/1000, 10)

woedf <- woe('loan_amnt_groups', train)
plotWoeNumeric('loan_amnt_groups ', woedf)
```

```{r loan_amnt adj}
adj_loan_amnt <- function(x){
  if (x < 8000) {
    return("[0,8000)")
  } else if(x < 11000){
    return("[8000, 11000)")
  } else if(x < 22000){
    return("[11000, 22000)")
  } else {
    return("22000+")
  }
}

train$loan_amnt_groups = sapply(train$loan_amnt, adj_loan_amnt)
woedf <- woe('loan_amnt_groups', train)
plotWoe('loan_amnt_groups', woedf)
```


## Interest rate

```{r interest_rate}
adj_int_rate <- function(x){
  if (x < 10) {
    return("[0,10)")
  } else if(x < 15){
    return("[10, 15)")
  } else if(x < 25){
    return("[15, 25)")
  } else {
    return("25+")
  }
}

train$int_rate_groups = sapply(train$int_rate, adj_int_rate)
woedf <- woe('int_rate_groups', train)
plotWoe('int_rate_groups', woedf)
```

## Past 2 year delinquencies

```{r}
adj_delinq_2yrs <- function(x){
  if(x == 0){
    return('0')
  } else if (x < 4){
    return('[1, 3]')
  } else {
    return('4+')
  }
}

train$delinq_2yrs_groups <- sapply(train$delinq_2yrs, adj_delinq_2yrs)
  
woedf <- woe('delinq_2yrs_groups', train)
plotWoe('delinq_2yrs_groups', woedf)
```

## DTI 

```{r}
# Droping the outliers
train$dti_c <- sapply(train$dti, capvalues, cap=40)

# Ploting the distribution
ggplot(data=train, aes(x=dti_c)) + geom_density(fill='blue')

# Cutting the annual income into 10 equal sized partitions
train$dti_groups = cut(train$dti_c, 7)

woedf <- woe('dti_groups', train)
plotWoeNumeric('dti', woedf)
```

```{r final_dti}
adj_dti <- function(x){
  if(x < 11){
    return("[0,11)")
  } else if (x < 17){
    return("[11,17)")
  } else if (x < 30){
    return("[17,30)")
  } else {
    return("30+")
  } 
}

train$dti_groups <- sapply(train$dti, adj_dti)

woedf <- woe('dti_groups', train)
plotWoe('Final dti splits', woedf)
```

## Public records

```{r public record}
adj_pub_rec <- function(x){
  if(x==0){
    return("0")
  } else {
    return(">0")
  }
}

train$pub_rec_groups = sapply(train$pub_rec, adj_pub_rec)

woedf <- woe('pub_rec_groups', train)
plotWoeNumeric('pub_rec_groups', woedf)
```

## Month differences between loan and cr line

```{r month_diff}
# Ploting the distribution
ggplot(data=train, aes(x=month_diff)) + geom_density(fill='blue')

# Cutting the annual income into 10 equal sized partitions
train$month_diff_groups = cut(train$month_diff , 10)

woedf <- woe('month_diff_groups', train)
plotWoeNumeric('month_diff_groups', woedf)
```

```{r final_month_diff}
adj_month_diff <- function(x){
  if(x<70){
    return("[0, 70)")
  } else if (x < 180){
    return("[70, 180)")
  } else if (x < 240) {
    return("[180, 240)")
  } else {
    return("240+")
  }
}

train$month_diff_groups = sapply(train$month_diff, adj_month_diff)

woedf <- woe('month_diff_groups', train)
plotWoe('month_diff_groups', woedf)
```

# Creating the final test and train frames 

```{r}
# Subseting the data
train <- d[issue_d_formated < zoo::as.yearmon('2018-01')]
test <- d[issue_d_formated >= zoo::as.yearmon('2018-01')]

# Creating the Y variables for modeling
train$Ypd <- sapply(train$loan_status, create_Ypd)
test$Ypd <- sapply(test$loan_status, create_Ypd)

train[, recovery_rate:=recoveries/funded_amnt]
train[, recovery_rate:=ifelse(recovery_rate>1, 1, recovery_rate)]

test[, recovery_rate:=recoveries/funded_amnt]
test[, recovery_rate:=ifelse(recovery_rate>1, 1, recovery_rate)]

train[, CCF:=(funded_amnt - total_rec_prncp)/funded_amnt]
test[, CCF:=(funded_amnt - total_rec_prncp)/funded_amnt]

# Train set adjustmetns
train$pub_recg <- sapply(train$pub_rec, adj_pub_rec)
train$annual_incg <- sapply(train$annual_inc, adj_income)
train$dtig <- sapply(train$dti, adj_dti)
train$month_diffg <- sapply(train$month_diff, adj_month_diff)
train$emp_lengthg <- sapply(train$emp_length, adj_emp_length)
train$int_rateg <- sapply(train$int_rate, adj_int_rate)
train$loan_amntg <- sapply(train$loan_amnt, adj_loan_amnt)
train$addr_stateg <- sapply(train$addr_state, adj_addr_state)
train$purposeg <- sapply(train$purpose, adj_purpose)

# Test set adjustments
test$pub_recg <- sapply(test$pub_rec, adj_pub_rec)
test$annual_incg <- sapply(test$annual_inc, adj_income)
test$dtig <- sapply(test$dti, adj_dti)
test$month_diffg <- sapply(test$month_diff, adj_month_diff)
test$emp_lengthg <- sapply(test$emp_length, adj_emp_length)
test$int_rateg <- sapply(test$int_rate, adj_int_rate)
test$loan_amntg <- sapply(test$loan_amnt, adj_loan_amnt)
test$addr_stateg <- sapply(test$addr_state, adj_addr_state)
test$purposeg <- sapply(test$purpose, adj_purpose)

# Feature list for PD model
featuresToUse = c(
  "issue_d_formated",
  "loan_status",
  "Ypd",
  "funded_amnt",
  "recovery_rate",
  "CCF",
  "term",
  "pub_recg",
  'annual_inc',
  "annual_incg",
  'dti',
  "dtig",
  'month_diff',
  "month_diffg",
  'emp_length',
  "emp_lengthg",
  'int_rate',
  "int_rateg",
  'loan_amnt',
  "loan_amntg",
  'grade',
  "addr_stateg",
  'purpose',
  'purposeg'
)

featuresMLPD <- c(
  "term",
  "pub_recg",
  "annual_incg",
  "dtig",
  "month_diffg",
  "emp_lengthg",
  "int_rateg",
  "loan_amntg",
  'grade',
  "addr_stateg",
  "purposeg"
)

# Beta regression features
featuresMLbeta <- c(
  "term",
  "pub_recg",
  "annual_inc",
  "dti",
  "month_diff",
  #"emp_lengthg",
  "int_rate",
  "loan_amnt",
  "grade",
  "addr_stateg",
  "purposeg"
)

# Subsetting 
train <- train[, ..featuresToUse]
test <- test[, ..featuresToUse]

# Saving the data for PD  
write.csv(na.omit(train), 'data/train.csv', row.names=FALSE)
write.csv(na.omit(test), 'data/test.csv', row.names=FALSE)

# Creating frames for charged off loans
loan_default_data_train = train[loan_status %in% c("Charged Off", "Does not meet the credit policy. Status:Charged Off")]
loan_default_data_test = test[loan_status %in% c("Charged Off", "Does not meet the credit policy. Status:Charged Off")]

# Saving the data for PD  
write.csv(na.omit(loan_default_data_train), 'data/train_defaults.csv', row.names=FALSE)
write.csv(na.omit(loan_default_data_test), 'data/test_defaults.csv', row.names=FALSE)

# Saving feature list
write.csv(featuresMLPD, "data/featuresMLPD.csv")
write.csv(featuresMLbeta, "data/featuresMLbeta.csv")
```
